High-Performance Language Primitives for Distributed Architectures using Asynchronous Events

### 2012-07-22 doc/asplos2013/micro.tex
edits to the micro benchmark section

In the single node case the latency of triggering an event is only 
329 nano-seconds.  For the multi-node cases, the triggering time is
between 3-4 micro-seconds which is the same as the cost of GASNet
active messages on Infiniband\cite{GASNET07}.  The gradual increase in latency with
node count is because each node must poll the Infiniband connection with all other nodes.
This demonstrates that the total cost of performing event triggers is in
the hardware and the underlying software stack.

@@

For small fan-in/out factors, throughput falls off initially going to two nodes as active
message become necessary, but starts increasing again at larger node counts.  Higher
fan-in/out factors require more messages and have lower throughput that increases with
node count.  In both cases the scaling of throughput is because the benchmark is
bound by the computational costs of active messages and not the bandwidth of the
interconnection network.  With more processors, more active messages can be processed
in parallel and performance improves.  
%Note we can further improve performance by increasing
%the number of threads per node for handling active messages (in these experiments there
%is one per node), but in real applications these threads would consume hardware cores and
%take away computational resources from the primary application.  This is only a viable
%option for memory or communication bound programs.

@@

\subsection{Lock Throughput}
\label{subsec:lockmicro}
Our lock micro-benchmark is designed to measure the maximum throughput of lock
acquisitions.  The benchmark is characterized by the number of nodes, the number
of locks initially created by each node, and the number of {\em chains} of
lock requests per node.  A chain is 1024 lock acquire and release operations that
are serialized by event dependences.  Each node creates the specified number of locks and
informs all other nodes of its lock handles.  All nodes create the specified number
of chains by randomly selecting from the set of all locks in the system.  All chains
are predicated on a user event.  The experiment measures the time from the start of
the user event to the completion of all chains.

Figure~\ref{fig:fixedlock} shows the lock grant throughput versus nodes in the system
for a fixed number of total chains in the system (256 total).  Each line corresponds
to a number of locks per node.  At larger node counts there will be more total locks
in the system.  At small node counts, small numbers of locks per node cause high
contention and lock unfairness allows for very high throughput as many requests are
handled locally before locks are migrated.  Large values have more total locks which
results in lower contention and lower throughput because more lock migrations are
required per satisfied request.  At larger node counts, there are enough locks
in the system to remove contention and throughput increases.

@@

To demonstrate that lock unfairness is also beneficial at larger node counts we show an alternative
cut of the experiment space.  In this case we fix the node count at 16 and plot
grant throughput as a function of chains per node for many different locks per node (also
total nodes in the system since node count is fixed).
The results can be seen in Figure~\ref{fig:fixednode}.  At very small chains per node
there is little contention and throughput is high.  At larger chains per node there is
more contention for locks and throughput drops because of higher lock migrations.  Note that 
as soon as chains per node is greater than total locks, the expected number of users per lock 
per node becomes more than one.  It is exactly at this point on each line that throughput
begins to improve again!  This is because as soon as there are multiple expected users on
a node, the unfairness property of the lock can amortize the cost of lock migration which
yields better throughput.

@@

\begin{itemize} \itemsep1pt \parskip0pt \parsep0pt
\item Single Reads/Writes - Each node takes a lock on the global physical region and then for each reduction
reads a values from the physical region, performs the reduction, and then write the result back.  The lock is then released.
\item Single Reductions - The reduction operation implementation performs each reduction individually
using an underlying active message implementation.
\item Localized Instance - Each node takes a lock on the the global physical region, copies it to a local physical
region, performs reductions to the localized region, copies the region back, and releases the lock.
\item Fold Instance - Each node creates a local reduction instance that folds reductions to the same element.
Reductions are applied locally and then the local reduction instance is merged back into the global region.
\item List Instance - Each node creates a local reduction instance that stores a list of reduction operations.
Reductions are applied locally and then the local reduction instance is merged back into the global region.
\end{itemize}

@@

The dense experiment demonstrates that using folding reduction regions perform 
best and scale with the number of nodes, achieving over a billion reductions
per second in the 16 node case.  List regions also scale, but perform about an
order of mangitude worse than folding regions in the dense case.  Not surprisingly
the other algorithms all have bottlenecks that result in poor performance.  The
sparse case is very similar with additional interesting behavior that at 16 nodes
the list regions begins performing better than the fold regions.  This shows that
list regions are better suited for scaling sparse reduction operations.

>>>>

In the single node case the latency of triggering an event is only 
329 nano-seconds.  For the multi-node cases, the triggering time is
between 3-4 micro-seconds which is the same as the cost of GASNet
active messages on Infiniband\cite{GASNET07}.  The gradual increase in latency with
node count is because each node must poll all the Infiniband connections for other nodes.
This cost of performing event triggers is in entirely in the hardware and the underlying software stack.

@@

For small fan-in/out factors, throughput falls off initially going to two nodes as active
message become necessary, but starts increasing again at larger node counts.  Higher
fan-in/out factors require more messages and have lower throughput that increases with
node count.  In both cases the scaling of throughput is because the benchmark is
bound by the computational costs of active messages and not the bandwidth of the
interconnection network.  With more processors, more active messages can be processed
in parallel and throughput improves.  
%Note we can further improve performance by increasing
%the number of threads per node for handling active messages (in these experiments there
%is one per node), but in real applications these threads would consume hardware cores and
%take away computational resources from the primary application.  This is only a viable
%option for memory or communication bound programs.

@@

\subsection{Lock Throughput}
\label{subsec:lockmicro}
Our lock micro-benchmark is designed to measure the maximum throughput of lock
grants.  The benchmark is characterized by the number of nodes, the number
of locks initially created by each node, and the number of {\em chains} of
lock requests per node.  A chain is 1024 lock acquire and release operations that
are serialized by event dependences.  Each node creates the specified number of locks and
informs all other nodes of its lock handles.  All nodes create the specified number
of chains by randomly selecting from the set of all locks in the system.  All chains
are predicated on a user event.  The experiment measures the time from the triggering of
the user event to the completion of all chains.

Figure~\ref{fig:fixedlock} shows the lock grant throughput versus nodes in the system
for 256 total chains in the system.  Each line corresponds
to a different number of initial locks per node.  At small node counts, 
small numbers of locks per node cause high
contention and lock unfairness allows for very high throughput as many requests are
handled locally before locks are migrated.   When there are more total locks there
is lower contention and lower throughput because the ratio of lock grants to migrated
locks decreases.  At larger node counts, there are enough locks
in the system to remove contention and throughput improves.

@@

To demonstrate that lock unfairness is also beneficial at larger node counts we show an alternative
cut of the experiment space.  In this case we fix the node count at 16 and plot
grant throughput as a function of chains per node for many different locks per node.
The results can be seen in Figure~\ref{fig:fixednode}.  At very small chains per node
there is little contention and throughput is high.  As the number of chains per node increases there is
more contention for locks and throughput drops because of more lock migrations.  Note that 
as soon as chains per node is greater than total locks, the expected number of users per lock 
per node becomes more than one.  It is exactly at this point on each line that throughput
begins to improve again.  This is because as soon as there are multiple expected users on
a node, the unfairness property of the lock can amortize the cost of lock migration which
yields better throughput.

@@

\begin{itemize} \itemsep1pt \parskip0pt \parsep0pt
\item Single Reads/Writes - Each node takes a lock on the global physical region and then for each reduction
reads a value from the physical region, performs the reduction, and then writes the result back.  The lock is then released.
\item Single Reductions - The reduction operation performs each reduction individually
using an underlying active message implementation.
\item Localized Instance - Each node takes a lock on the the global physical region, copies it to a local physical
region, performs reductions to the localized region, copies the region back, and releases the lock.
\item Fold Instance - Each node creates a local reduction instance that folds reductions to the same element.
Reductions are applied locally and then the local reduction instance is merged back into the global region.
\item List Instance - Each node creates a local reduction instance that stores a list of reduction operations.
Reductions are applied locally and then the local reduction instance is merged back into the global region.
\end{itemize}

@@

The dense experiment demonstrates that folding reduction regions perform 
best and scale with the number of nodes, achieving over a billion reductions
per second in the 16 node case.  List regions also scale, but perform about an
order of mangitude worse than folding regions in the dense case.  Not surprisingly
the other algorithms all have bottlenecks that result in poor performance.  The
sparse case is very similar with the additional interesting occurence that at 16 nodes
list regions begin performing better than fold regions.  This shows that
list regions are better suited for scaling sparse reduction operations.
